---
title: "Simulation Example on Structural Equation Modeling"
author: "PSYC 621"
date: "4/29/2019"
output: 
  html_document: 
    toc: true
---

\newcommand{\var}{\textrm{Var}}
\newcommand{\bv}[1]{\boldsymbol{\mathbf{#1}}}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = ">#")
comma <- function(x, digits. = 2L) format(x, digits = digits., big.mark = ",")
```

# Simulation Example on Structural Equation Modeling (SEM)

```{r load_pkg, message=FALSE}
# Load required packages
library(tidyverse)
theme_set(theme_bw())
```


## Simulate Multivariate Data

In SEM, when multivariate normality is assumed, one can either generate data 
directly using matrix algebra, or generate the latent variables first before 
generating the observed variables. The first method is faster, but the second
method is more general and can be applied to situations like categorical data or
multilevel data. Therefore, in this note we'll use the second method. 

Let's do a latent growth model (LGM) similar to the one in the note "Simulating
Multilevel Data." Here is the model in a latent growth model representation:
$$\begin{bmatrix}
    y_{0i} \\
    y_{1i} \\
    y_{2i} \\
    y_{3i}
  \end{bmatrix} = 
  \bv \Lambda 
  \begin{bmatrix}
    \eta_{1i} \\
    \eta_{2i}
  \end{bmatrix} + 
  \begin{bmatrix}
    e_{0i} \\
    e_{1i} \\
    e_{2i} \\
    e_{3i}
  \end{bmatrix}$$
where $y_{0i}, \ldots, y_{3i}$ are the outcome values for person $i$ from time 
0 to time 3, $\eta_{1i}$ is the specific intercept for person $i$, $\eta_{2i}$ 
is the specific slope for person $i$, and $e_{0i}, \ldots, e_{3i}$ are the 
within-person level error term. The distributional assumptions are 

\begin{align*}
  \begin{bmatrix}
    \eta_{1i} \\
    \eta_{2i}
  \end{bmatrix} & \sim 
  \mathcal{N}\left(\begin{bmatrix}
                \alpha_1 \\
                \alpha_2
              \end{bmatrix}, 
              \begin{bmatrix}
                \phi_{11} & \phi_{21} \\
                \phi_{21} & \phi_{22}
              \end{bmatrix}\right) \\
  \begin{bmatrix}
    e_{0i} \\
    e_{1i} \\
    e_{2i} \\
    e_{3i}
  \end{bmatrix} & \sim
  \mathcal{N}\left(\begin{bmatrix}
                0 \\
                0 \\
                0 \\
                0
              \end{bmatrix}, 
              \begin{bmatrix}
                \theta_{11} & 0 & 0 & 0 \\
                0 & \theta_{22} & 0 & 0 \\
                0 & 0 & \theta_{33} & 0 \\
                0 & 0 & 0 & \theta_{44} \\
              \end{bmatrix}\right)
\end{align*}

A path diagram is shown below:

```{r, message=FALSE}
growth_model <- "i =~ 1*y1 + 1*y2 + 1*y3 + 1*y4
                 s =~ 0*y1 + 1*y2 + 2*y3 + 3*y4
                 i ~~ 1 * i
                 s ~~ 0.2 * s + 0.1 * i
                 y1 ~~ 0.5 * y1
                 y2 ~~ 0.5 * y2
                 y3 ~~ 0.5 * y3
                 y4 ~~ 0.5 * y4
                 i ~ 1 * 1
                 s ~ 0.5 * 1"
library(semPlot)
semPaths(semPlotModel_lavaanModel(growth_model))
```

### Using R

Here's the code for generating the data:

```{r}
library(mnormt)
set.seed(123)
# Define sample size
N <- 100
# Define the Fixed Parameters
alpha <- c(1, 0.5)  # latent means
Phi <- matrix(c(1, 0.1, 
                0.1, 0.2), nrow = 2)  # latent variances/covariances
Lambda <- cbind(c(1, 1, 1, 1), 
                c(0, 1, 2, 3))  # factor loadings
Theta <- diag(0.5, nrow = 4)
# Generate latent factor scores
eta <- rmnorm(N, mean = alpha, varcov = Phi)
# Generate residuals:
e <- rmnorm(N, varcov = Theta)
# Compute outcome scores
y <- tcrossprod(eta, Lambda) + e
# Make it a data frame
colnames(y) <- paste0("y", 1:4)
df <- as.data.frame(y)
```

If you want to check whether the simulated data is correct, generate with a large sample size, and check the means and covariances:

```{r}
N_test <- 1e5
eta_test <- rmnorm(N_test, varcov = Phi)
colMeans(eta_test)
cov(eta_test)
e_test <- rmnorm(N_test, varcov = Theta)
colMeans(e_test)
cov(e_test)
```

It's handy to wrap this as a function:

```{r}
gen_lgm_data <- function(N, alpha, Phi, Lambda, Theta) {
  # Generate latent factor scores
  eta <- rmnorm(N, mean = alpha, varcov = Phi)
  # Generate residuals:
  e <- rmnorm(N, varcov = Theta)
  # Compute outcome scores
  y <- tcrossprod(eta, Lambda) + e
  colnames(y) <- paste0("y", 1:4)
  # Make it a data frame
  as.data.frame(y)
}
# Test it:
set.seed(123)
gen_lgm_data(100, 
             alpha = c(1, 0.5), 
             Phi = matrix(c(1, 0.1, 
                            0.1, 0.2), nrow = 2), 
             Lambda = cbind(c(1, 1, 1, 1), 
                            c(0, 1, 2, 3)), 
             Theta = diag(0.5, nrow = 4)) %>% 
  head() # shows only the first six cases
```

Indeed, you get the same data!

### Other methods for generating SEM data

Many SEM software or packages have capability in generating data with input of
an SEM model. For example, in R, you can call Mplus using the 
`MplusAutomation` package and use their `MONTECARLO` routine. In R, you can 
generate SEM data using the `lavaan` package with the `simulateData()` 
function, like the following example:

```{r}
# Using a previously defined SEM model:
lavaan::simulateData(growth_model) %>% 
  head() # shows only the first six cases
```

Personally, however, I prefer directly simulating data in R because 

- it forces you to specify everything in the model in the way you want.
Especially in Mplus there are a lot of hidden default settings that may mess up
with your simulation; 
- it makes the process of generating data more transparent;
- it helps you learn the math behind the model;
- it is more flexible as you can specify any distributional assumptions or 
models not supported by the SEM packages. 

## Analyzing the Simulated Data

In R, for running SEM models, the most common options are `lavaan`, `OpenMx`,
and Mplus (via `MplusAutomation`). When possible, I'll stick to `lavaan` to
avoid jumping between programs, so let's analyze the simulated data twice, first
with the true model and second with a misspecified model where the random slope
term is omitted (i.e., the variance of `s` is constrained to zero). 

```{r}
library(lavaan)
# True model
m1 <- 'i =~ 1*y1 + 1*y2 + 1*y3 + 1*y4
       s =~ 0*y1 + 1*y2 + 2*y3 + 3*y4
       i ~~ s'
# Model without random slopes
m2 <- 'i =~ 1*y1 + 1*y2 + 1*y3 + 1*y4
       s =~ 0*y1 + 1*y2 + 2*y3 + 3*y4
       s ~~ 0*i + 0*s'
m1_fit <- growth(m1, data = df)
m2_fit <- growth(m2, data = df)
```

Here's some examples of things you can extract (without showing all the results)

```{r, eval=FALSE}
fitMeasures(m1_fit)  # return more than 30 fit indices
# Specific fit indices
fitMeasures(m1_fit, c("chisq", "df", "pvalue", "cfi", "rmsea"))
coef(m1_fit)  # parameter estimates
vcov(m1_fit)  # asymptotic covariance matrix of parameters
# Asymptotic standard errors of latent mean of s
sqrt(diag(vcov(m1_fit))["s~1"])
parameterEstimates(m1_fit, standardized = FALSE)  # coefficients, SE, CI
modificationIndices(m1_fit, minimum.value = 3.84)  # modification indices
# Bootstrapped fit index
bootstrapLavaan(m1_fit, R = 1000L, type = "bollen.stine",
                FUN = fitMeasures, fit.measures = c("cfi"))
```

## Full Example of a Small Scale Simulation

In a methodological experiment with Monte Carlo simulation, one usually
generates millions of data sets across tens or hundreds of carefully chosen
conditions. As an example, here is a small scale simulation study on LGM. The
two goals are: (a) to understand the bias on the mean of slopes and its standard
error estimates, and (b) to illustrate the difference between estimated and
empirical standard error.

For simplicity, I'll only choose three __designed factors__ (i.e., manipulated
independent variables), namely sample size, variance of the random slope in the
data generating model, and the mean of the slopes. The design factors are 
summarized here:

- Sample size (_N_): 50, 100, 200
- Variance of slopes ($\phi_{22}$): 0.1, 0.5 (i.e., 1/10 and 1/2 of the intercept
variance)
- Mean of slopes ($\alpha_2$): 1, 0.5

Therefore, it's a 3 $\times$ 2 $\times$ 2 factorial design. Now let's walk
through each component. 

### Fixed Values for the Study

In the study I decided to run 500 replications for each condition. I can
estimate the simulation error to see whether it's enough.

```{r}
set.seed(515)  # set the seed for reproducibility
NREP <- 500  # number of replications
# Fixed parameters (all caps)
ALPHA1 <- 1  # latent mean of intercepts
PHI11 <- 1  # intercept variance
LAMBDA <- cbind(c(1, 1, 1, 1), 
                c(0, 1, 2, 3))  # factor loadings
THETA <- diag(0.5, nrow = 4)  # residual variances
```

We should also define everything that will not change across replications and
across conditions, including syntax for running growth models, functions to 
generate data, etc. 

```{r}
# Function for generating data:
gen_lgm_data <- function(N, alpha, Phi, Lambda, Theta) {
  # Generate latent factor scores
  eta <- rmnorm(N, mean = alpha, varcov = Phi)
  # Generate residuals:
  e <- rmnorm(N, varcov = Theta)
  # Compute outcome scores
  y <- tcrossprod(eta, Lambda) + e
  colnames(y) <- paste0("y", 1:4)
  # Make it a data frame
  as.data.frame(y)
}
# lavaan syntax
# True model
m1 <- 'i =~ 1*y1 + 1*y2 + 1*y3 + 1*y4
       s =~ 0*y1 + 1*y2 + 2*y3 + 3*y4
       i ~~ s'
# Model without random slopes
m2 <- 'i =~ 1*y1 + 1*y2 + 1*y3 + 1*y4
       s =~ 0*y1 + 1*y2 + 2*y3 + 3*y4
       s ~~ 0*i + 0*s'
```

### Keep track of simulation conditions

I'll create a data frame to store the design factors:

```{r}
# Design factors:
DESIGNFACTOR <- expand.grid(
  N = c(50, 100, 200), 
  phi22 = c(0.1, 0.5), 
  alpha2 = c(1, 0.5)
)
# Add condition number:
DESIGNFACTOR <- rowid_to_column(DESIGNFACTOR, "cond")
DESIGNFACTOR
```

An additional benefit is that I only need to loop over one dimension (condition
number) instead of writing three loops, one for each factor. 

### Writing a function to conduct the simulation

While this step is not necessary, it's recommended to wrap your steps of 
generating data and extracting results into one big function, so that the code
is easier to read, like the `runsim()` function below:

```{r}
runsim <- function(to_run,  # conditions to run
                   nrep,  # number of replications
                   alpha1 = ALPHA1,  # latent mean of intercepts
                   phi11 = PHI11,  # intercept variance
                   Lambda = LAMBDA,  # factor loadings
                   Theta = THETA,  # residual variances
                   designfactors = DESIGNFACTOR) {
  # Extract design parameters for the given condition
  N <- designfactors[to_run, "N"]
  phi22 <- designfactors[to_run, "phi22"]
  alpha2 <- designfactors[to_run, "alpha2"]
  # Put the values back to the matrix
  alpha <- c(alpha1, alpha2)
  Phi <- matrix(c(phi11, phi22 / 2,
                  phi22 / 2, phi22), nrow = 2)
  # Initialize place holders for the results
  rep_result <- vector("list", nrep)
  for (i in seq_len(nrep)) {
    # Generate data
    df <- gen_lgm_data(N, alpha, Phi, Lambda, Theta)
    # Run model 1
    m1_fit <- growth(m1, data = df)
    # Run model 2
    m2_fit <- growth(m2, data = df)
    # Save results 
    rep_result[[i]] <- list(m1_fit = m1_fit, 
                            m2_fit = m2_fit)
  }
  # Return results
  return(rep_result)
}
```

We can do a trial run:

```{r}
# Run condition 1 with 2 replications
sim_results <- runsim(1, 2)
# Examine the results:
sim_results[[1]]
```

### Running the simulation

This step is easy. You can use a for loop, or the `purrr::map()` function

```{r, eval=FALSE}
sim_results <- 
  map(seq_len(nrow(DESIGNFACTOR)), 
      ~ runsim(.x, nrep = NREP))
```


### Extract the results

Right now I'm saving everything for each step. I'll need a function to extract
the relevant output, which are

+ Estimated coefficients of $\alpha_2$
+ Estimated standard errors of $\alpha_2$

It's usually helpful to start extract results on one replication first:

```{r, eval=FALSE}
# Try on one replication
with(sim_results[[1]][[1]], 
     tibble(coef(m1_fit)["s~1"], 
            sqrt(vcov(m1_fit)["s~1", "s~1"]), 
            coef(m2_fit)["s~1"], 
            sqrt(vcov(m2_fit)["s~1", "s~1"])))
# Now, wrap it as a function:
extract_coef <- function(res) {
  out <- with(res, 
              tibble(coef(m1_fit)["s~1"], 
                     sqrt(vcov(m1_fit)["s~1", "s~1"]), 
                     coef(m2_fit)["s~1"], 
                     sqrt(vcov(m2_fit)["s~1", "s~1"])))
  # Add names:
  names(out) <- c("m1_est", "m1_se", "m2_est", "m2_se")
  out
}
extract_coef(sim_results[[1]][[1]])
```

Now, we can do it on all replications and all conditions:

```{r, eval=FALSE}
library(tidyverse)
sim_outdata <- sim_results %>% 
  # Iterate across conditions
  map_dfr(
    # Iterate across replications
    ~ map_dfr(.x, extract_coef, .id = "rep"), 
    .id = "cond") %>% 
  # Make `cond` a integer variable
  mutate(rep = as.integer(rep), 
         cond = as.integer(cond)) %>% 
  # Add design factors
  left_join(DESIGNFACTOR)
# Save the results
write_csv(sim_outdata, "example_sem_results.csv")
```

```{r, include=FALSE}
sim_outdata <- read_csv("example_sem_results.csv")
```


### Summarize the Results

Finally, let's summarize the results in terms of the standardized bias and the relative standard error bias. 

```{r}
sim_outdata <- sim_outdata %>% 
  # Make it long format
  gather("var", "val", m1_est:m2_se) %>% 
  separate(var, c("model", "var")) %>% 
  spread(var, val)
sim_sum <- sim_outdata %>% 
  # Summarize results by conditions
  group_by(model, alpha2, phi22, N) %>% 
  summarise(ave_est = mean(est), 
            emp_sd = sd(est), 
            ave_se = mean(se)) %>% 
  # Compute standardized bias and relative SE bias
  mutate(bias = ave_est - alpha2, 
         bias_mcse = emp_sd / sqrt(NREP), 
         std_bias = bias / emp_sd, 
         rse_bias = (ave_se - emp_sd) / emp_sd)
```

With relatively small number of conditions, one can present the results in a 
table (and it's handy in R):

```{r}
sim_sum %>% 
  select(model:N, bias, bias_mcse, std_bias, rse_bias) %>% 
  knitr::kable(digits = 3L)
```

It is, however, recommended you try to plot the results, both for exploratory
purpose and for better presentation of the results. 

```{r}
# Summarize estimates
sim_outdata %>% 
  ggplot(aes(x = factor(N), y = est, color = model)) + 
  geom_boxplot() + 
  geom_hline(aes(yintercept = alpha2)) + 
  facet_grid(alpha2 ~ phi22)
# Standardized bias
sim_sum %>% 
  ggplot(aes(x = factor(N), y = std_bias, color = model)) + 
  geom_line(aes(group = model)) + 
  geom_hline(aes(yintercept = 0)) + 
  facet_grid(alpha2 ~ phi22)
# Relative SE bias
sim_sum %>% 
  ggplot(aes(x = factor(N), y = rse_bias, color = model)) + 
  geom_line(aes(group = model)) + 
  geom_hline(aes(yintercept = 0)) + 
  facet_grid(alpha2 ~ phi22)
```

## Exercise

1. From the simulation results, evaluate the relative efficiency of the 
estimated average slope (i.e., $\alpha_2$) under model 2 relative to that under
model 1. 

